---
title: "Capstone: Bellabeat Data Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Document description 

This document outlines all of the steps taken to clean and transform Bellabeats's raw datasets to prepare the data for the next stage of analysis. The dataset description can be located [here.](https://www.kaggle.com/datasets/arashnic/fitbit)

## 2. Loading necessary libraries and datasets

### 2.1 Loading Packages and Libraries
```{r}
#Installing the packages
install.packages('tidyverse') #used for data cleaning
install.packages('lubridate') #for date & time formats
install.packages('janitor') #used for cleaning data
install.packages('formattable')
install.packages('patchwork') #main title for subplots
```

```{r}
#Loading Libraries
library(tidyverse) #drop_na()
library(lubridate)
library(janitor) #clean_names()
library(dplyr)
library(ggplot2)
library(formattable)
library(patchwork)
```

### 2.2 Loading Datasets

```{r load datasets}
daily_activity <- read.csv("/cloud/project/dailyActivity_merged.csv")
sleep_day <- read.csv("/cloud/project/sleepDay_merged.csv")
weight_log_info <- read.csv("/cloud/project/weightLogInfo_merged.csv")
hourly_intensity <- read.csv("/cloud/project//hourlyIntensities_merged.csv")
```

### 2.3 Summary of dataframes
```{r explore dataframes}
head(daily_activity) 
head(sleep_day)
head(weight_log_info) 
head(hourly_intensity) 
```

## 3. Exploring and Cleaning the data

### 3.1 Changing case from camel case to snake case
```{r changing case}
daily_activity <- clean_names(daily_activity)
sleep_day <- clean_names(sleep_day)
weight_log_info <- clean_names(weight_log_info)
hourly_intensity <- clean_names(hourly_intensity)
```

### 3.2 Checking for duplicates
```{r}
sum(duplicated(daily_activity))
sum(duplicated(sleep_day))
sum(duplicated(weight_log_info))
sum(duplicated(hourly_intensity))
```

### 3.3 Deleting duplicates
```{r}
# Before deleting duplicates
nrow(sleep_day)

# Command to clear duplicates
sleep_day <- sleep_day %>%
  distinct() %>%
  drop_na()

# After deleting duplicates
nrow(sleep_day)
```

### 3.4 Changing datatype for date from char to date datatype
```{r Changing datatype for date from char to date datatype}
daily_activity$activity_date <- as.Date(daily_activity$activity_date,'%m/%d/%y')
sleep_day$sleep_day <- as.Date(sleep_day$sleep_day,'%m/%d/%y')
weight_log_info$date <- as.Date(weight_log_info$date,'%m/%d/%y')
```

### 3.5 Renaming column activity_date and sleep_day to date
```{r renaming column activity_date and sleep_day to date}
daily_activity <- daily_activity %>%
  rename(date = activity_date)

sleep_day <- sleep_day %>%
  rename(date = sleep_day)
```

### 3.6 Converting minutes to hours for total_minutes_asleep
```{r}
#total hours asleep
sleep_day$total_hours_asleep = round((sleep_day$total_minutes_asleep)/60, digits = 2)

head(sleep_day)
```

### 3.7 Dropping total_minutes_asleep in sleep_day dataset
```{r}
sleep_day = subset(sleep_day, select = -c(total_minutes_asleep))
```

### 3.8 Formatting String datatype to Boolean datatype
```{r}
weight_log_info$is_manual_report <- as.logical(weight_log_info$is_manual_report)

head(weight_log_info)
```

### 3.9 Formatting activity_hour to date time datatype in hourly intensity dataframe
```{r}
hourly_intensity$activity_hour=as.POSIXct(hourly_intensity$activity_hour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
```

### 3.10 Separating activity_hour into date and time columns for hourly intensity dataframe
```{r}
hourly_intensity$time <- format(hourly_intensity$activity_hour, format = "%H:%M:%S")
hourly_intensity$date <- format(hourly_intensity$activity_hour, format = "%m/%d/%y")
```

### 3.11 Creating a new column for weekdays
```{r}
# Adding a new column for weekdays
daily_activity$day_of_week <- wday(daily_activity$date, label = T, abbr = T)

# After adding the weekday column
head(daily_activity)
```

### 3.12 Merging datasets
```{r}
daily_activity_sleep <- merge(daily_activity, sleep_day, by=c ("id", "date"))
daily_activity_weight <- merge(daily_activity, weight_log_info, by=c ("id", "date"))

head(daily_activity_sleep)
head(daily_activity_weight)
```

## 4 Analysis

### 4.1 Total Steps Vs Sleep
```{r}
ggplot(data = daily_activity_sleep) +
  aes(x=total_steps, y = total_hours_asleep) +
  geom_point(color = '#004c6d') +
  geom_smooth() + 
  theme(panel.border = element_rect(colour = "black", fill=NA)) +
  labs(x = 'Total Steps', y = 'Total Hours Asleep', 
       title = 'Total Steps vs Total Hours Asleep')

ggsave('Total Steps vs Total Minutes Asleep.png')
```

## 4.2 Total Steps Vs Weekday
```{r}
#options(scipen=) will remove any scientific notations
options(scipen = 999)

ggplot(data=daily_activity_sleep) + 
  geom_col(mapping = aes(x=day_of_week,y=total_steps, fill=day_of_week)) + 
  theme(panel.border = element_rect(colour = "black", fill=NA)) +
  scale_fill_brewer(palette="Set2") +
  theme(legend.position="none") +
  labs(x = 'Day of the week', y = 'Total Steps', 
  title = 'Total no of steps taken in a week')

ggsave("Total no of steps taken in a week.png")
```

## 4.3 Sleep Distribution
```{r}
sleep_log <- sleep_day%>%
  mutate(sleep_log = case_when( 
     total_hours_asleep < 7 ~ "Less than 7 hours",
     total_hours_asleep >= 7 &  total_hours_asleep < 8 ~ "7 to 8 hours", 
     total_hours_asleep >= 8 ~ "More than 8 hours"
  ))

head(sleep_log)
```

#percentage 
```{r}
sleep_percent <- sleep_log%>%
  group_by(sleep_log) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(sleep_log) %>%
  summarise(sleep_percent = total / totals) %>%
  mutate(percent = scales::percent(sleep_percent))
```
#Formattable use to turn chr to percent
```{r}
sleep_percent <- sleep_percent %>% mutate(percent=percent(sleep_percent))

head(sleep_percent)
```

```{r}
ggplot(data=sleep_percent) + 
  geom_col(mapping = aes(x=sleep_log,y=percent, fill=sleep_log)) + 
  theme(panel.border = element_rect(colour = "black", fill=NA)) +
  scale_fill_brewer(palette="Set2") +
  theme(legend.position="none") +
  labs(x = 'Total Hours Asleep', y = 'Percent', title = 'Sleep Distribution')

ggsave("Sleep Distribution.png")
```

## 4.4 Total Steps Vs Calories
```{r}
daily_activity_sleep %>% 
  group_by(total_steps, calories) %>% 
  ggplot(aes(x = total_steps, y = calories, color = calories)) +
  geom_point() +
  geom_smooth() + 
  theme(legend.position = c(.9, .2),
        legend.spacing.y = unit(1, "mm"), 
        panel.border = element_rect(colour = "black", fill=NA),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black")) +
  labs(title = 'Calories Burned by Total Steps Taken',
       y = 'Calories',
       x = 'Total Steps')

ggsave("Calories Burned by Total Steps Taken.png")
```

## 4.5 BMI Distribution
```{r}
weight_log_bmi <- weight_log_info %>%
  mutate(bmi_log = case_when( 
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 24 ~ "Normal Weight", 
    bmi >= 24 & bmi < 29.9 ~ "Overweight", 
    bmi >= 30 ~ "Obesity",
  ))

head(weight_log_bmi)
```

### 4.5.1 Distinct users
```{r}
weight_log_users <- weight_log_bmi %>% distinct(id, .keep_all = TRUE)
weight_log_bmi %>% group_by(id) %>% filter(row_number() == 1)
```

### 4.5.2 Percentage 
```{r}
weight_bmi_percent <- weight_log_users %>%
  group_by(bmi_log) %>%
  summarise(total = n()) %>%
  mutate(totals = sum(total)) %>%
  group_by(bmi_log) %>%
  summarise(bmi_percent = total / totals) %>%
  mutate(percent = scales::percent(bmi_percent))
```
#Formattable use to turn chr to percent
```{r}
weight_bmi_percent <- weight_bmi_percent %>% mutate(percent=percent(bmi_percent))

head(weight_bmi_percent)
```

### 4.5.3 BMI Distribution
```{r}
ggplot(weight_bmi_percent,aes(fill=bmi_log,y = percent,x = "")) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette="Set2") +
  theme_void() +
  geom_text(aes(label = percent, x=1),position = position_stack(vjust = 0.5)) +
  theme(plot.title = element_text(hjust = 0.5, vjust = -5, size = 15, family="TT Times New Roman", face = "bold")) +
  labs(title = "BMI Distribution") +
  guides(fill = guide_legend(title = "BMI Distribution"))

ggsave("BMI Distribution.png")
```

## 4.6 Weight Vs Total Steps
```{r}
ggplot(data = daily_activity_weight) +
  aes(x=weight_kg, y=total_steps) +
  geom_point(color = '#004c6d') +
  geom_smooth() + 
  theme(panel.border = element_rect(colour = "black", fill=NA)) +
  labs( x = 'Weight in kg', y = 'Total Steps Taken',
       title = 'Total Steps Taken Vs Weight in Kg')

ggsave("weight vs total steps.png")
```

## 4.7 User Type Distribution
```{r}
daily_average <- daily_activity_sleep %>%
  group_by(id) %>%
  summarise (mean_total_steps = mean(total_steps),
             mean_total_calories = mean(calories),
             mean_total_sleep = mean(total_hours_asleep))

head(daily_average)
```
#According to 10000 steps,

#Sedentary: Less than 5,000 steps per day
#Low active: Between 5,000 to 7,499 steps per day
#Somewhat active: Between 7,500 to 9,999 steps per day
#Active: More than 10,000 steps per day
#Highly active: More than 12,500

```{r}
activity_user_type <- daily_average %>%
  mutate(activity_user_type = case_when( 
    mean_total_steps < 5000 ~ "Sedentary",
    mean_total_steps >= 5000 & mean_total_steps < 7500 ~ "Low active", 
    mean_total_steps >= 7500 & mean_total_steps < 10000 ~ "Somewhat active", 
    mean_total_steps >= 10000 & mean_total_steps < 12500 ~ "Active",
    mean_total_steps >= 12500 ~ "Highly active"
  ))

head(activity_user_type)
```
#Now that we have put the users in different active levels based on the mean total number of steps taken. Let's find out the percentages for the same.

```{r}
user_type <- activity_user_type %>%
    group_by(activity_user_type) %>%
    summarise(total = n()) %>%
    mutate(totals = sum(total)) %>%
    group_by(activity_user_type) %>%
    summarise(user_percent = total / totals) %>%
    mutate(percent = scales::percent(user_percent))
```
#Formattable use to turn chr to percent
```{r}
user_type <- user_type %>% mutate(percent=percent(user_percent))

head(user_type)
```

#Let's plot a pie chart to see the different active levels and which of each level is highest and which is the lowest.

```{r}
ggplot(user_type,aes(fill=activity_user_type,y = percent,x = "")) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette="Set2") +
  theme_void() +
  geom_text(aes(label = percent, x=1),position = position_stack(vjust = 0.5)) +
  theme(plot.title = element_text(hjust = 0.5, vjust = -5, size = 15, family="TT Times New Roman", face = "bold")) +
  labs(title = "User Type Distribution") +
  guides(fill = guide_legend(title = "User Type Distribution"))
  
ggsave("User Type Distribution.png")
```

## 4.8 Active Hours

# Let's plot a graph to see the peak hours

```{r}
hourly_intensity <- hourly_intensity %>%
  group_by(time) %>%
  drop_na() %>%
  summarise(mean_hourly_intensity = mean(total_intensity))
```

```{r}
ggplot(data = hourly_intensity) +
  aes(x=time, y = mean_hourly_intensity) +
  geom_histogram(stat = "identity", fill='#004c6d') +
  theme(panel.border = element_rect(colour = "black", fill=NA)) +  
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = 'Hour', y = 'Total Intensity', 
       title = 'Total Intensity per hour')
       
ggsave("Active Hours.png")
```
